---
title: "Hillary 2016 Contributions"
author: "Philip Seifi"
date: "July 20, 2015"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
#
# DATA SOURCE: http://fec.gov/disclosurep/PDownload.do
# FORMAT: ftp://ftp.fec.gov/FEC/Presidential_Map/2016/DATA_DICTIONARIES/CONTRIBUTOR_FORMAT.txt
# --------------------------------------------------------------- 
# 
# CMTE_ID			COMMITTEE ID			            	S - skipped
# CAND_ID			CANDIDATE ID			            	S - skipped
# CAND_NM			CANDIDATE NAME			            	S - skipped
# CONTBR_NM	    	CONTRIBUTOR NAME		            	S
# CONTBR_CITY		CONTRIBUTOR CITY		            	S
# CONTBR_ST		    CONTRIBUTOR STATE		            	S
# CONTBR_ZIP		CONTRIBUTOR ZIP CODE                    S
# CONTBR_EMPLOYER	CONTRIBUTOR EMPLOYER		        	S
# CONTBR_OCCUPATION	CONTRIBUTOR OCCUPATION		        	S
# CONTB_RECEIPT_AMT	CONTRIBUTION RECEIPT AMOUNT	        	N
# CONTB_RECEIPT_DT	CONTRIBUTION RECEIPT DATE	        	D	
# RECEIPT_DESC		RECEIPT DESCRIPTION		            	S
# MEMO_CD			MEMO CODE				                S - skipped
# MEMO_TEXT	    	MEMO TEXT			                	S
# FORM_TP			FORM TYPE			                	S - skipped
# FILE_NUM		    FILE NUMBER				                N - skipped
# TRAN_ID			TRANSACTION ID				            S - skipped
# ELECTION_TP		ELECTION TYPE/PRIMARY GENERAL INDICATOR S - skipped
#

# Load contribution data from CSV
hrc_in = read.csv('hillary-contribs.csv',
                  header=TRUE,
                  colClasses=c(rep('NULL', 3),
                               rep(NA, 9),
                               'NULL', NA,
                               rep('NULL', 4)),
                  strip.white=TRUE)

str(hrc_in)
summary(hrc_in)
```

The highest contributing state is California. The highest contributing city is New York.

The median contribution amount is $250.

I am omitting all contributions above the $2,700 limit (Source: http://www.fec.gov/pages/fecrecord/2015/february/contriblimits20152016.shtml) as they break the Federal Election Campaign Act and thus have or will be refunded.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Omitt contributions above the $2,700 FEC limit
hrc_in = filter(hrc_in, contb_receipt_amt <= 2700)
```

To get us started, let's look at a density plot on contribution amounts, omitting refunds (negative contributions).

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Omitt refunds (negative contributions)
hrc_in_positive = filter(hrc_in, hrc_in$contb_receipt_amt > 0)
# Plot density of contribution amounts
ggplot(aes(x=contb_receipt_amt), data=hrc_in_positive) +
    geom_density()
```

We can observe a U-shaped curve indicating that most contributions fall in one of two groups: small contributions of around $200, and large contributions at the maximum FEC limit of $2,700. There is also a small bump around the $1,000 mark, possibly an artificial, round contribution limit people decide upon in their minds.

My hunch is that the two groups contain very different demographics, so let's go on and explore contributions by occupation.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get top 10 occupations by number of contributions
num_by_occup = hrc_in %>%
                filter(contbr_occupation != '',
                        contbr_occupation != 'INFORMATION REQUESTED') %>%
                group_by(contbr_occupation) %>%
                tally() %>%
                arrange(desc(n)) %>%
                top_n(10)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 occupations by number of contributions
ggplot(aes(x=contbr_occupation, y=n), data=num_by_occup) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

Most of the contributors so far are from retired supporters, closely followed by attorneys, and in a distant third, consultants.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get top 10 occupations by total value of contributions
val_by_occup = hrc_in %>%
                filter(contbr_occupation != '',
                       contbr_occupation != 'INFORMATION REQUESTED') %>%
                group_by(contbr_occupation) %>%
                tally(contb_receipt_amt) %>%
                arrange(desc(n)) %>%
                top_n(10)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 occupations by total value of contributions
ggplot(aes(x=contbr_occupation, y=n), data=val_by_occup) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

The picture is different if we plot total contribution amounts rather than the number of contributions. Now it's the attorneys who take the first place, followed by retired, homemakers and consultants. This suggests that consultants and the retired make lower average contributions than attorneys and homemakers. Let's confirm this assumption...

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get top 10 occupations by median value of contributions
median_val_by_occup = hrc_in %>%
                        filter(contbr_occupation != '',
                               contbr_occupation != 'INFORMATION REQUESTED') %>%
                        group_by(contbr_occupation) %>%
                        summarize(n=n(), m=median(contb_receipt_amt)) %>%
                        arrange(desc(n)) %>%
                        top_n(10, n)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 occupations by median value of contributions
ggplot(aes(x=contbr_occupation, y=m), data=median_val_by_occup) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

Indeed, it appears that although their total contributions are significant, the median contributions of the retired and consultants are relatively low. Physicians make surprisingly minor contributions given their presumably above-average income. Homemakers, on the other hand, have some of the highest median contributions among top 10 contibutors in this campaign, up there with CEOs and Presidents.

There's clearly a lot of interesting insight to be found here, and a box plot might give us a better overview of this data...

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get contributions by occupation
contbs_top_occups = filter(hrc_in,
                           contbr_occupation %in%
                               num_by_occup[['contbr_occupation']])
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Box plot contributions by occupation
ggplot(aes(x=contbr_occupation, y=contb_receipt_amt), data=contbs_top_occups) +
    geom_boxplot(outlier.colour = "red", outlier.size = 2) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

Interesting, but refunds (negative contributions) aren't particularly helpful in this case and make the plot less readable. Let's omitt them and try again...

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get contributions by occupation, omitting refunds
contbs_top_occups = filter(hrc_in,
                           contbr_occupation %in%
                               num_by_occup[['contbr_occupation']],
                           contb_receipt_amt > 0)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot contributions by occupation, omitting refunds
ggplot(aes(x=contbr_occupation, y=contb_receipt_amt), data=contbs_top_occups) +
    geom_boxplot(outlier.colour = "red", outlier.size = 2) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

We can see that the average contributions are indeed quite low for the retired, unemployed, professors and physicians. There is significant variation in these groups, however, with some supporters making comparable contributions to those from the other occupations.

We can also observe a lot of variation among attorneys and presidents, some of them contributing up to 3x more than their respective medians. This is likely indicative of the U-shaped [distribution of salaries among attorneys](http://qph.is.quoracdn.net/main-qimg-0a0d8f37efe16a83e4f1208aea3b1988?convert_to_webp=true).

Finally, there is equally some negative variation among CEOs. My presumption is that the lower contributions are made by CEOs of startups and other SMEs, who might not have disposable income comparable to that of corporate CEOs classified under the same group.

The differences among occupations are certainly interesting, but some of the largest donors in political campaigns tend to be corporations. Let's take a look contribution amounts by employer. I will omitt the retired, and the unemployed, because they vastly outweigh individual companies and organizations, and we've already had a look at these categories earlier.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get top 10 employers by total value of contributions
val_by_empl = hrc_in %>%
                filter(contbr_employer != '',
                       contbr_employer != 'INFORMATION REQUESTED',
                       contbr_employer != 'N/A',
                       contbr_employer != 'RETIRED',
                       contbr_employer != 'NOT EMPLOYED') %>%
                group_by(contbr_employer) %>%
                tally(contb_receipt_amt) %>%
                arrange(desc(n)) %>%
                top_n(10)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 employers by total value of contributions
ggplot(aes(x=contbr_employer, y=n), data=val_by_empl) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

The self-employed are clearly, by far, the largest contributors to Hillary Clinton's campaign. This is surprising, given that the self-employed traditionally vote GOP. Of course, without analysis of all contributions in this electoral cycle, it is impossible to tell whether an even larger number of self-employed Americans contribute to Republican candidates, as one would expect, given that Republicans are roughly 50% more likely to be self-employed (Fried, pp. 104-5, 125.)

Let's redraw the same graph omitting the self-employed to have a better look at the individual institutions. I manually colour-coded each of the top employers by their industry.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get top 10 employers by total value of contributions, omitt self-employed
val_by_empl_noself = val_by_empl %>%
                        filter(contbr_employer != 'SELF-EMPLOYED')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 employers by total value of contributions, omitt self-employed
industry <- c(rep("LAW",4),
              rep("FINANCE",1),
              rep("TALENT",1),
              rep("UNIVERSITY",1),
              rep("LAW",2))
ggplot(aes(x=contbr_employer, y=n, fill=industry), data=val_by_empl_noself) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

This chart makes it especially clear where much of the money comes from. The largest donor is Morgan & Morgan, a consumer protection and personal injury law firm. Most of the other donors are also legal firms, with the exception of one talent agency, one financial institution and one Ivy League university.

Given that some of the top donors in the 2012 Obama campaign (Source: https://www.opensecrets.org/pres12/) were major universities, I would have expected to see more purple in the list. I'm also surprised to see Morgan Stanley, one of the top donors in the 2012 Mitt Romney campaign (ibid.). Of course, here again, it is impossible to say whether Morgan Stanley doesn't have an equal, or even larger stake in the campaigns of GOP candidates without analysis of the entire dataset.

It would be most interesting to analyze contributions by gender, given that the Hillary 2016 campaign actively targets female voters. The FEC dataset does not contain contributor genders, but this information can be prediceted based on first names of the contributors, using the `gender` package and the Social Security Administration gender data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(gender)
# Strip last and middle names
hrc_in$contbr_fnm = sub(" .*", "", sub(".*, ", "", hrc_in$contbr_nm))
# Sample 5,000 entries from the data set (gender prediction is a very slow process)
set.seed(20022012)
hrc_in_samp <- hrc_in[sample(1:nrow(hrc_in), 5000), ]
hrc_in_samp$gender = gender(as.character(hrc_in_samp$contbr_fnm),
                            method = "ssa",
                            years = 1970) %>%
                        do.call(rbind.data.frame, .)
hrc_in_samp$gender = hrc_in_samp$gender$gender
# Omitt refunds (negative contributions)
hrc_in_samp = filter(hrc_in_samp, contb_receipt_amt > 0)
```

First let's look at a density distribution by gender.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot density of contribution amounts
ggplot(aes(x=contb_receipt_amt, color=gender), data=hrc_in_samp) +
    geom_density()
```

It appears that women tend to make lower contributions that men. This is likely the effect of lower salaries and greater number of homemakers among women (i.g, stay-at-home wifes).

Let's test this assumption by generating the same plot, but omitting homemakers from the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Omitt homemakers
hrc_in_samp_no_home = filter(hrc_in_samp,
                             contbr_occupation != 'HOMEMAKER',
                             contbr_occupation != '',
                             contbr_occupation != 'INFORMATION REQUESTED',
                             contbr_occupation != 'N/A')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot density of contribution amounts
ggplot(aes(x=contb_receipt_amt, color=gender), data=hrc_in_samp_no_home) +
    geom_density()
```

My assumptiom was clearly wrong. The new subset in fact shows an even lower number of high-amount contributions among women than before.

What about total contributions by gender? A couple of pie charts can be a good way to visualize this... 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get number of contributions by gender
num_by_gender = hrc_in_samp %>%
                group_by(gender) %>%
                tally()
num_by_gender = na.omit(num_by_gender)
# Get total value of contributions by gender
val_by_gender = hrc_in_samp %>%
                group_by(gender) %>%
                tally(contb_receipt_amt)
val_by_gender = na.omit(val_by_gender)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(gridExtra)
# Plot pie chart of contribution number by gender
num_pie = ggplot(aes(x="", y=n, fill=gender), data=num_by_gender) +
    geom_bar(width = 1, stat = "identity") +
    coord_polar("y", start=0)
# Plot pie chart of contribution total value by gender
val_pie = ggplot(aes(x="", y=n, fill=gender), data=val_by_gender) +
    geom_bar(width = 1, stat = "identity") +
    coord_polar("y", start=0)
grid.arrange(num_pie, val_pie, ncol=2)
```

The first pie chart is the number of contributions by gender. The second, the total value by gender. It appears that although the number of contributions is larger for female supporters, the total value is comparable.

This confirms the findings from the density plot that women tend to make smaller contributions than men, but also alligns with Hillary's appeal to female voters.

Next, let's make a choropleth map of states by total contributions to see where the money flows from...

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(RColorBrewer)
library(choroplethr)
library(choroplethrMaps)
library(maptools)

# Get top 10 states by total value of contributions
val_by_state = hrc_in %>%
                filter(contbr_st != '',
                       contbr_st != 'INFORMATION REQUESTED',
                       contbr_st != 'N/A') %>%
                group_by(contbr_st) %>%
                summarize(value = sum(contb_receipt_amt)) %>%
                arrange(desc(value))

# Expand state abbreviations into a new column with full, lowercase state names
val_by_state$region = tolower(state.name[match(val_by_state$contbr_st,
                                               state.abb)])
val_by_state = na.omit(val_by_state)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot choropleth map of states by total value of contributions
state_choropleth(val_by_state)
```

Again, some surprising results worth investigating. Why is Texas, a predominantly Republican state, one of the top sources of contribution for Hillary?

Let's start with a density plot on contribution amounts, omitting refunds (negative contributions), coming from the state of Texas.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Generate a subset of contributions coming from Texas
hrc_in_texas = filter(hrc_in, contbr_st == 'TX', contb_receipt_amt > 0)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot density of Texas contribution amounts
ggplot(aes(x=contb_receipt_amt), data=hrc_in_texas) +
    geom_density()
```

Let's redraw the occupation bar charts using a subset of contributions coming from Texas...

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get top 10 occupations from Texas by number of contributions
texas_num_by_occup = hrc_in_texas %>%
                filter(contbr_occupation != '',
                        contbr_occupation != 'INFORMATION REQUESTED') %>%
                group_by(contbr_occupation) %>%
                tally() %>%
                arrange(desc(n)) %>%
                top_n(10)
```

Nothing suprising to see here. The distribution is similar to that of contributions nationwide, with the exception of a less noticeable bump at the $1,000 level and a slightly lower peak at the $2,700 FEC limit.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 occupations from Texas by number of contributions
ggplot(aes(x=contbr_occupation, y=n), data=texas_num_by_occup) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Get top 10 occupations from Texas by total value of contributions
texas_val_by_occup = hrc_in_texas %>%
                filter(contbr_occupation != '',
                       contbr_occupation != 'INFORMATION REQUESTED') %>%
                group_by(contbr_occupation) %>%
                tally(contb_receipt_amt) %>%
                arrange(desc(n)) %>%
                top_n(10)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 occupations from Texas by total value of contributions
ggplot(aes(x=contbr_occupation, y=n), data=texas_val_by_occup) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

Attorneys appear to be an even more dominant occupation in Texas than in the rest of the country, both in number and total amount of contributions.

Let's see if a plot of employers confirms this finding.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Fix name inconsistencies that severely skewed the results
hrc_in_texas$contbr_employer = gsub("INC.", "INC", hrc_in_texas$contbr_employer)
hrc_in_texas$contbr_employer = gsub(", ", " ", hrc_in_texas$contbr_employer)

# Get top 10 employers from Texas by total value of contributions
texas_val_by_empl = hrc_in_texas %>%
                        filter(contbr_employer != '',
                               contbr_employer != 'INFORMATION REQUESTED',
                               contbr_employer != 'N/A',
                               contbr_employer != 'RETIRED',
                               contbr_employer != 'NOT EMPLOYED',
                               contbr_employer != 'SELF-EMPLOYED') %>%
                        group_by(contbr_employer) %>%
                        tally(contb_receipt_amt) %>%
                        arrange(desc(n)) %>%
                        top_n(10)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot top 10 employers from Texas by total value of contributions
industry <- c(rep("ENERGY",1),
              rep("LAW",1),
              rep("IT",1),
              rep("LAW",2),
              rep("HEALTH",1),
              rep("LAW",1),
              rep("HEALTH",1),
              rep("TELECOM",1),
              rep("LAW",1))
ggplot(aes(x=contbr_employer, y=n, fill=industry), data=texas_val_by_empl) +
    geom_bar(stat='identity') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4))
```

Surprisingly, after all, law firms are not the dominants sources of contributions in Texas, compared to the national totals. Could this be caused by a smaller average size of law firms in the state?

It is also interesting to note that the top contributor is an energy company, a sector which is not represented among the top 10 nationwide.

Let's redraw the choropleth, but this time zoom onto the state of Texas and investigate individual ZIP codes.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(choroplethrZip)
#Convert ZIP+4 codes to 5-digit ZIP codes
hrc_in$contbr_zip = strtrim(hrc_in$contbr_zip, 5)
val_by_zip = hrc_in %>%
                filter(contbr_st == 'TX') %>%
                group_by(contbr_zip) %>%
                summarize(value = sum(contb_receipt_amt)) %>%
                arrange(desc(value))
val_by_zip$region = val_by_zip$contbr_zip
val_by_zip = na.omit(val_by_zip)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
texas_zip_choro = ZipChoropleth$new(val_by_zip)
texas_zip_choro$ggplot_scale = scale_fill_brewer(palette=1, drop=FALSE)
texas_zip_choro$set_zoom_zip(state_zoom="texas", county_zoom=NULL, msa_zoom=NULL, zip_zoom=NULL)
texas_zip_choro$render()
```

The more fine-grained choropleth of contributions from Texas shows that most contributions come from a few specific regions of the state. This is not surprisingly, but any further conclusions require a better knowledge of the demographics of Texas than I possess, and ideally data on political affiliations of the individual regions.

Contributions from Texas are certainly worth looking into, but now, let's take a look at how contributions change throughout the campaign.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Convert dates from strings to the Date datatype
hrc_in$contb_receipt_dt = as.Date(hrc_in$contb_receipt_dt, "%d/%m/%Y")
# Omitt refunds
hrc_in_positive = filter(hrc_in, contb_receipt_amt > 0,
                         !is.na(contb_receipt_dt))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Time series plot contributions from April to July 2015
ggplot(aes(x=contb_receipt_dt, y=contb_receipt_amt), data=hrc_in_positive) +
    geom_point() +
    geom_smooth(formula = contb_receipt_amt ~ contb_receipt_dt)
```

There's some overplotting in the lower levels of the contributions amount gradient. The data also gives the false perception that all contributions are made at the same time of each day. Let's add a little bit of transparency and position jitter to avoid falling to false conclusions.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Time series plot contributions from April to July 2015
ggplot(aes(x=contb_receipt_dt, y=contb_receipt_amt), data=hrc_in_positive) +
    geom_point(alpha=0.3, position='jitter') +
    geom_smooth(formula = contb_receipt_amt ~ contb_receipt_dt)
```

We can see that outide of the occasional outlier, the contribution amounts appear to be remarkably consistent. A an increase of small contributions can be observed in the last few months. This could be seasonal, or a natural evolution as the electoral cycle progresses. 

Of note are also frequent contributions at the $250, $500, $1,000 levels, likely a psychological effect of people choosing round values for their donations.

Let's add some colours... red for Republican states, blue for Democrat states, and green for swing states.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#
# Add political spectrum colour by state
#
#       Colour guide:
#           0 for blue states
#           1 for red states
#           0.5 for swing (green) states
#
# Based on https://en.wikipedia.org/wiki/Red_states_and_blue_states#/media/File:Red_state,_blue_state.svg
# (summary of results of the 2000, 2004, 2008, and 2012 presidential elections)
#
state_spectrum = data.frame(contbr_st = c("AK","AL","AR","AZ","CA","CO","CT","DC","DE","FL","GA","GU","HI","IA","ID", "IL","IN","KS","KY","LA","MA","MD","ME","MH","MI","MN","MO","MS","MT","NC","ND","NE","NH","NJ","NM","NV","NY","OH","OK","OR","PA","PR","PW","RI","SC","SD","TN","TX","UT","VA","VI","VT","WA","WI","WV","WY"), spectrum = c(1,1,1,1,0,0.5,0,0,0,0.5,1,0,0,0,1,0,1,1,1,1,0,0,0,0,0,0,1,1,1,1,1,1,0,0,0,0.5,0,0.5,1,0,0,0,0,0,1,1,1,1,1,0.5,0,0,0,0,1,1))
hrc_in_positive = merge(hrc_in_positive, state_spectrum, by="contbr_st")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Time series plot contributions by state political affiliation
ggplot(aes(x=contb_receipt_dt,
           y=contb_receipt_amt,
           color=I(as.factor(spectrum))),
       data=hrc_in_positive) +
    geom_point(alpha=0.3, position='jitter') +
    geom_smooth(formula = contb_receipt_amt ~ contb_receipt_dt) +
    scale_colour_manual(values=c("blue", "green", "red"))
```

Although the median contributions are consistently higher from Democrat states, the difference isn't very large. Of note are the last two weeks of May, when median Republican and swing state contributions were higher than those from Democrat states.

##Final plots and summary

Exploratory Data Analysis of the Hillary Rodham Clinton 2016 campaign from April to July 2016 identified several avenues for further investigation.

###1. Contributions by occupation (nationwide)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Box plot contributions by top 10 occupations
ggplot(aes(x=contbr_occupation, y=contb_receipt_amt), data=contbs_top_occups) +
    geom_boxplot(outlier.colour = "red", outlier.size = 2) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .4)) +
    ggtitle('Contributions by occupation (nationwide)') +
    labs(x="", y="Amount in US$")
```

The average contributions to the campaign differ substantially across occupations.

The retired, unemployed, professors and physicians all have very low median contributions, with some outliers in the $2000-2700 range.

CEOs, homemakers and presidents have a median equal to the contribution limit. Attorneys, lawyers, and consultants do not make high median contributions but there is very significant variance. This is likely because some of these contributors are employed by major law and consulting firms who may even endorse and cover their contributions, whereas others are part of a smaller partnership and cover the contribution out of pocket.

The variance may also be indicative of the U-shaped [distribution of salaries in the legal professions](http://qph.is.quoracdn.net/main-qimg-0a0d8f37efe16a83e4f1208aea3b1988?convert_to_webp=true).

###3. Contributions by state
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot choropleth map of states by total value of contributions
state_choropleth(val_by_state,
                 title = "Contributions by state",
                 legend = "Amount in US$",
                 num_colors = 1)
```

The choropleth map of total amount of contributions by state is largely unsurprising, with one exception.

Why is Texas, a predominantly Republican state, one of the top sources of contributions for Hillary Clinton?

I tried to investigate this question by exploring contributions by employer using a Texas subset of the data, but discovered no easy explanation.

It may be worthwhile to explore the Texas subset further, at a more granular level, mapping contributions by ZIP code.

###4. Contribution time series
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Time series plot contributions by state political affiliation
ggplot(aes(x=contb_receipt_dt,
           y=contb_receipt_amt,
           color=I(as.factor(spectrum))), data=hrc_in_positive) +
    geom_point(alpha=0.3, position='jitter') +
    geom_smooth(formula = contb_receipt_amt ~ contb_receipt_dt) +
    scale_colour_manual(values=c("blue", "green", "red"),
                        labels=c("Democrat","Swing state","Republican"),
                        guide = guide_legend(title = 'Source region')) +
    ggtitle('Contributions since April 1st 2015') +
    labs(x="Date", y="Amount in US$")
```

Plotting contributions from April 1st till July 1st, gives us yet another perspective on the data. 

Contributions appear to be remarkably consistent. A an increase in small contributions (< $250) can be observed in the last few months. This could be seasonal, or a natural evolution as the electoral cycle progresses.

Although the median contributions are consistently higher from Democrat states, the difference isn’t very large. Of note are the last two weeks of May, when median Republican and swing state contributions were higher than those from Democrat states.

##Reflection
The Hillary 2016 campaign contributions dataset contains over 38,000 entrie from April till July 2015. Although the elections are still far away, analysis of existing data can give us some indication of most important contribution sources, and allows us to predict the demographics Hillary Clinton should approach as her fundraising progresses.

I started by understanding the individual variables by studying the [official dataset format](ftp://ftp.fec.gov/FEC/Presidential_Map/2016/DATA_DICTIONARIES/CONTRIBUTOR_FORMAT.txt), then conducting basic descriptive statistical analysis.

To begin with, I chose to omitt all entries above the [$2700 contribution limit](http://www.fec.gov/pages/fecrecord/2015/february/contriblimits20152016.shtml) as they break the [Federal Election Campaign Act](http://www.fec.gov/law/feca/feca.pdf) and thus were or will be refunded.

I then used these findings and my domain expertise to explore contribution numbers and amounts across a number of variables including occupation, employer, and state.

Some findings where not surprising. Contributions to Hillary's campaign tend to come from California and the city of New York, they tend to be made by the retired, and law firms are some of the major employers behind these contributors.

Other findings were more surprising, however:

* Physicians, who are [among the best paid in the country](http://www.bls.gov/oes/current/oes_nat.htm),  make very small contributions compared to other occupations.

* Texas, a [predominantly Republican state](https://en.wikipedia.org/wiki/Politics_of_Texas), is among the top 10 sources of contributions to Hillary Clinton's campaign.

* The self-employed are clearly, by far, the largest contributors to Hillary Clinton's campaign. This is surprising, given that Republicans are roughly 50% more likely to be self-employed (Fried, pp. 104–5, 125.). Of course, without analysis of all contributions in this electoral cycle, it is impossible to tell whether an even larger number of self-employed Americans contribute to GOP candidates.

It would be most interesting to explore this dataset further by combining it with data from Hillary Clinton's 2008 presidential campaign. Have Hillary's contribution sources changed since 8 years ago? Could we predict current and future contributors using data from previous elections? These are all important questions that could be studied using a combined dataset.